R Notebook

### 2019-03-13 analysis/INLA/test-INLA-sensor.Rmd
running INLA with multiple years

######################################
# Start Bayesian analysis
# One more time the model:

# Movement_ij      ~ Gamma(mu_ik, r)
# E(Movement_ij)   = mu_ij
# var(Movement_ij) = mu_ij^2 / r 
# mu_ij = exp(Intercept + fRepro + Annual Trend + Seasonal Trend + a_i)

# Below we define a model with a factor fRepro, a long term
# trend for Year, and seasonal trend for DayInYear, and
# a random intercept for bear.

# Each trend is modelled as a random walk trend.
# So each of them has a sigma.

```{r}
fit_inla_sensor <- function(dat, response = "temperature_C_avg",
                     n_knots = 50,
                     family = "binomial",
                     max_edge = c(20, 100),
                     kmeans = FALSE,
                     plot = FALSE,
                     fit_model = TRUE,
                     extend = list(n = 8, offset = -0.1),
                     offset = c(5, 25),
                     cutoff = 10,
                     include_depth = TRUE,
                     verbose = FALSE,
                     debug = FALSE,
                     trials = 10) {
  d <- dat
  coords <- as.matrix(unique(d[, c("X", "Y")]))

  # use kmeans() to calculate centers:
  if (kmeans) {
    km <- stats::kmeans(x = coords, centers = n_knots)
    mesh <- INLA::inla.mesh.create(km$centers,
      extend = extend
    )
  } else {
    bnd <- INLA::inla.nonconvex.hull(coords)
    mesh <- INLA::inla.mesh.2d(
      offset = offset,
      boundary = bnd,
      max.edge = max_edge,
      cutoff = cutoff
    )
  }

  message("INLA max_edge = ", "c(", max_edge[[1]], ", ", max_edge[[2]], ")")
  if (plot) {
    graphics::plot(mesh, asp = 1)
    graphics::points(coords, col = "red")
  }

  spde <- INLA::inla.spde2.matern(mesh, alpha = 3 / 2)
  dat <- data.frame(
    y = d[, response, drop = TRUE],
    xcoo = d$X,
    ycoo = d$Y
  )

  dat$intercept <- 1

  if (include_depth) {
    dat$depth <- d$depth_scaled
    dat$depth2 <- d$depth_scaled2
  }

  # construct index for ar1 model:
  iset <- INLA::inla.spde.make.index(
    "i2D",
    n.spde = mesh$n
  )

  # make the covariates:
  X.1 <- dat[, -c(1:3), drop = FALSE]
  covar_names <- colnames(X.1)
  XX.list <- as.list(X.1)
  effect.list <- list()
  effect.list[[1]] <- c(iset)
  for (Z in seq_len(ncol(X.1)))
    effect.list[[Z + 1]] <- XX.list[[Z]]
  names(effect.list) <- c("1", covar_names)

  A <- INLA::inla.spde.make.A(
    mesh = mesh,
    loc = cbind(dat$xcoo, dat$ycoo)
  )
  A.list <- list()
  A.list[[1]] <- A
  for (Z in seq_len(ncol(X.1)))
    A.list[[Z + 1]] <- 1

  # make projection points stack:
  sdat <- INLA::inla.stack(
    tag = "stdata",
    data = list(y = dat$y),
    A = A.list,
    effects = effect.list
  )

  formula <- as.formula(paste(
    "y ~ -1 +",
    paste(covar_names, collapse = "+"),
    "+ f(i2D, model=spde)"
  ))

  if (fit_model) {
    try_fitting <- TRUE
    trial_i <- 1
    while (try_fitting && trial_i <= trials) {
      model <- try(INLA::inla(formula,
        family = family,
        data = INLA::inla.stack.data(sdat),
        control.predictor = list(
          compute = TRUE,
          A = INLA::inla.stack.A(sdat)
        ),
        control.fixed = list(
          mean = 0, prec = 1 / (5^2),
          mean.intercept = 0, prec.intercept = 1 / (20^2)
        ),
        control.compute = list(config = TRUE),
        verbose = verbose,
        debug = debug,
        keep = FALSE
      ))
      trial_i <- trial_i + 1
      if (class(model) != "try-error") {
        try_fitting <- FALSE
      } else {
        # if (trial_i > 5) browser()
        message("INLA error. Trying again.")
      }
    }
    if (class(model) == "try-error") {
      model <- NA
    }
  } else {
    model <- NA
  }
  list(
    model = model, mesh = mesh, spde = spde, data = dat,
    formula = formula, iset = iset
  )
}
```

```{r}
obj <- fit_inla_sensor(temp_dat, response = "temperature_C_avg",
                       n_knots = 50,
                       family = "gaussian",
                       max_edge = c(20, 100),
                       kmeans = FALSE,
                       plot = FALSE,
                       fit_model = TRUE,
                       extend = list(n = 8, offset = -0.1),
                       offset = c(5, 25),
                       cutoff = 10,
                       include_depth = TRUE,
                       verbose = FALSE,
                       debug = FALSE,
                       trials = 2)

pg <- gfplot:::make_prediction_grid(filter(temp_dat, year %in% c(2016, 2017)), survey ="SYN QCS")$grid

cell_area <- gfplot:::make_prediction_grid(filter(temp_dat, year %in% c(2016, 2017)), survey = "SYN QCS")$cell_area

pred_temp <- predict_inla_sens(obj=obj, pred_grid = pg , samples = 100L,
                         include_depth = TRUE)
```

@@

```{r posterior-draws}
pg_one_year <- filter(pg, year == min(pg$year))
predicted <- predict_inla_sens(obj = obj, pred_grid = pg_one_year, samples = 100, include_depth = TRUE)
```

@@

samples = 10L
pred_grid <- pg

predict_inla_sens <- function(obj, pred_grid, samples = 100L,
                         include_depth = TRUE) {

  mesh <- obj$mesh
  model <- obj$model
  iset <- obj$iset

@@

  # get indices of various effects:
  latent_names <- rownames(inla.mcmc[[1]]$latent)
  re_indx <- grep("^i2D", latent_names)
  if (include_depth) {
    depth_fe_indx1 <- grep("^depth$", latent_names)
    depth_fe_indx2 <- grep("^depth2$", latent_names)
  }
  intercept_indx <- grep("^intercept$", latent_names)

@@

  # projections will be stored as an array:
  projected_latent_grid <-
    array(0, dim = c(nrow(grid_locs), samples))
  proj_matrix <- INLA::inla.spde.make.A(mesh, loc = as.matrix(grid_locs))

  # do MCMC projections:
  indx <- which(iset$i2D.group == 1)
  for (i in seq_len(samples)) {

    if (include_depth) {
      depth_effects <-
        pred_grid[, "depth_scaled", drop = TRUE] * inla.mcmc[[i]]$latent[depth_fe_indx1] +
        pred_grid[, "depth_scaled2", drop = TRUE] * inla.mcmc[[i]]$latent[depth_fe_indx2]
    } else {
      depth_effects <- 0
    }

    projected_latent_grid[, i] <-
      as.numeric(proj_matrix %*% inla.mcmc[[i]]$latent[re_indx][indx]) +
      inla.mcmc[[i]]$latent[intercept_indx] +
      depth_effects
  }

  projected_latent_grid
}

@@

Calculate probability * Gamma median values on the grid:

```{r prediction-grid-calcs}
pg_one_year <- filter(pg, year == min(pg$year))
attributes(predicted)
# out <- stats::plogis(p_bin) * exp(p_pos)
out_med <- apply(predicted, c(1), median)
# out_bin <- stats::plogis(p_bin)
# out_bin <- apply(out_bin, c(1, 3), median)
dim(predicted)

pg_all <- bind_rows(replicate(1, pg_one_year, simplify = FALSE))

out_long <- reshape2::melt(out_med)
pg_all$pred <- out_long$value
```

@@

make_plot <- function(dat, fill = "sqrt(pred)") {
  g <- ggplot(dat, aes(X * 10, Y * 10)) +
    geom_raster(aes_string(fill = fill)) +
    scale_fill_viridis_c(option = "C") +
    #facet_wrap(~ year) +
    geom_polygon(
      data = coast, aes_string(x = "X", y = "Y", group = "PID"),
      fill = "grey80"
    ) +
    geom_point(
      data = filter(d, present == 1),
      aes(x = X * 10, y = Y * 10, size = density), inherit.aes = FALSE,
      pch = 21, col = "white", alpha = 0.3
    ) +
    geom_point(
      data = filter(d, present == 0),
      aes(x = X * 10, y = Y * 10), inherit.aes = FALSE,
      pch = 4, col = "white", alpha = 0.2
    )

@@

make_plot(pg_all, "pred") + 
  scale_fill_gradient2(
    midpoint = median(pg_all$pred),
    low = scales::muted("blue"), 
    mid = "white",
    high = scales::muted("red"))

>>>>

######################################
 Start Bayesian analysis
 One more time the model:
 Movement_ij      ~ Gamma(mu_ik, r)
 E(Movement_ij)   = mu_ij
 var(Movement_ij) = mu_ij^2 / r
 mu_ij = exp(Intercept + fRepro + Annual Trend + Seasonal Trend + a_i)
 Below we define a model with a factor fRepro, a long term
 trend for Year, and seasonal trend for DayInYear, and
 a random intercept for bear.
 Each trend is modelled as a random walk trend.
 So each of them has a sigma.

d <- temp_dat
response = "temperature_C_avg"
n_knots = 50
family = "gaussian"
plot = TRUE
fit_model = TRUE
extend = list(n = 8, offset = -0.1)
RangeGuess = 100 
offset = c(RangeGuess/5, RangeGuess)
max_edge = c(RangeGuess/5,RangeGuess)
cutoff = (RangeGuess/5)/5
prior_range = c(RangeGuess, 0.01) # P(range < 0.05) = 0.01
prior_sigma = c(0.5, 0.5)
kmeans = FALSE    
verbose = FALSE
debug = FALSE
trials = 2
covariates = c("depth_scaled", "depth_scaled2")
time = "year"

fit_inla_sensor <- function(d, response = "temperature_C_avg",
                     n_knots = 50,
                     family = "binomial",
                     plot = FALSE,
                     fit_model = TRUE,
                     extend = list(n = 8, offset = -0.1),
                     kmeans = FALSE,                   
                     RangeGuess = 100, 
                     offset = c(RangeGuess/5, RangeGuess),                 
                     max_edge = c(RangeGuess/5,RangeGuess),
                     cutoff = (RangeGuess/5)/5,
                     prior_range = c(RangeGuess, 0.01), # P(range < 0.05) = 0.01
                     prior_sigma = c(0.5, 0.5),
                     time = "year",
                     covariates = NULL,
                     verbose = FALSE,
                     debug = FALSE,
                     trials = 10) {

  coords <- as.matrix(unique(d[, c("X", "Y")]))

  # use kmeans() to calculate centers:
  if (kmeans) {
    km <- stats::kmeans(x = coords, centers = n_knots)
    mesh <- INLA::inla.mesh.create(km$centers,
      extend = extend
    )
  } else {
    bnd <- INLA::inla.nonconvex.hull(coords)

    mesh <- INLA::inla.mesh.2d(
      offset = offset,
      boundary = bnd,
      max.edge = max_edge,
      cutoff = cutoff
    )
  }
  message("INLA max_edge = ", "c(", max_edge[[1]], ", ", max_edge[[2]], "), n = ", mesh$n)

  if (plot) {
    graphics::plot(mesh, asp = 1)
    graphics::points(coords, col = "red")
  }

  # create data.frame including response ("y"), coordinates("xcoo", "ycoo"), covariates, and "intercept"
  dat <- data.frame(
    y = d[, response, drop = TRUE],
    xcoo = d$X,
    ycoo = d$Y, 
    raw_time = d[, time, drop = TRUE],
    select(d, covariates)
  )

  # make a design matrix where the first year is the intercept:
  YO  <- order(dat$raw_time)
  dat <- dat[YO,]

  time_k <- as.numeric(as.factor(dat$raw_time))

  #time_k <- dat$raw_time - min(dat$raw_time) + 1
  k <- max(time_k)
  time_lab <- paste0("Y", seq(1, k))
  dat[time_lab] <- 0
  dat[, time_lab[1]] <- 1
  for (j in seq_along(time_lab)) {
    dat[time_k == j, time_lab[j]] <- 1
  }
  #names(dat)[names(dat)=="Y1"] <- "intercept"

  # create an object for a Matérn model
  spde <- INLA::inla.spde2.matern(mesh = mesh, alpha = 3 / 2) #, # WHAT DOES ALPHS DO?????
  #                                prior.range = prior_range, 
  #                                prior.sigma = prior_sigma)

  # construct index for ar1 model ---- DON'T QUITE UNDERSTAND THIS
  iset <- INLA::inla.spde.make.index(
    name = "spatial.field",
    n.spde = mesh$n,
    n.group = k
  )

  # isolate list of covariates:
  X.1              <- dat[, -c(1:4), drop = FALSE] # remove "y", "xcoo", "ycoo", "real_time"
  covar_names      <- colnames(X.1) # list covariates
  XX.list          <- as.list(X.1) # turn data.frame to a list
  effect.list      <- list()
  effect.list[[1]] <- c(iset)

 for (Z in seq_len(ncol(X.1)))
    effect.list[[Z + 1]] <- XX.list[[Z]]

  names(effect.list) <- c("1", covar_names)

 # projection matrix to map to observed points
 A <- INLA::inla.spde.make.A(
    mesh = mesh,
    loc = cbind(dat$xcoo, dat$ycoo),
    group = time_k
  )
  A.list <- list()
  A.list[[1]] <- A

  for (Z in seq_len(ncol(X.1)))
    A.list[[Z + 1]] <- 1

  # make projection points stack:
  sdat <- INLA::inla.stack(tag = "stdata",
                           data = list(y = dat$y),
                           A = A.list,
                           effects = effect.list
  )

  formula <- as.formula(paste(
    "y ~ -1 +",
    paste(covar_names, collapse = "+"),
    "+ f(spatial.field, model=spde)"
  ))

  if (fit_model) {
    try_fitting <- TRUE
    trial_i <- 1
    while (try_fitting && trial_i <= trials) {
      model <- try(INLA::inla(formula,
                   family = family,
                   data = INLA::inla.stack.data(sdat),
                   control.predictor = list(compute = TRUE,
                                            A = INLA::inla.stack.A(sdat)),
                   control.fixed     = list(mean = 0, 
                                            prec = 1 / (5^2),
                                            mean.intercept = 0, 
                                            prec.intercept = 1 / (20^2)),
                   control.compute = list(config = TRUE),
                   verbose = verbose,
                   debug = debug,
                   keep = FALSE
                 )
      )
      trial_i <- trial_i + 1
      if (class(model) != "try-error") {
        try_fitting <- FALSE
      } else {
        # if (trial_i > 5) browser()
        message("INLA error. Trying again.")
      }
    }
    if (class(model) == "try-error") {
      model <- NA
    }
  } else {
    model <- NA
  }
  list(
    model = model, mesh = mesh, spde = spde, data = dat,
    formula = formula, iset = iset
  )
}
```

```{r}
obj <- fit_inla_sensor(temp_dat, response = "temperature_C_avg",
                       n_knots = 50,
                       family = "gaussian",
                       plot = TRUE,
                       fit_model = TRUE,
                       extend = list(n = 8, offset = -0.1),
                       RangeGuess = 100, 
                       covariates = c("depth_scaled", "depth_scaled2"),
                       verbose = FALSE,
                       debug = FALSE,
                       trials = 2)

pg <- gfplot:::make_prediction_grid(temp_dat, survey ="SYN QCS")$grid #filter(temp_dat, year %in% c(2016, 2017))
cell_area <- gfplot:::make_prediction_grid(filter(temp_dat, year %in% c(2016, 2017)), survey = "SYN QCS")$cell_area
```

@@

@@

# samples = 10L
# pred_grid <- pg_one_year

predict_inla_sens <- function(obj, pred_grid, 
                                   samples = 100L) {
  mesh <- obj$mesh
  model <- obj$model
  iset <- obj$iset

@@

  # get indices of various effects:
  latent_names   <- rownames(inla.mcmc[[1]]$latent)
  yr_fe_indx     <- grep("^Y[0-9]+$", latent_names)
  re_indx        <- grep("^spatial.field", latent_names)
  depth_fe_indx1 <- grep("^depth_scaled$", latent_names)
  depth_fe_indx2 <- grep("^depth_scaled2$", latent_names)
  #intercept_indx <- grep("^intercept$", latent_names)

@@

  # define empty arrays to store projections:
  latent_grid   <- array(0, dim = c(nrow(grid_locs), samples, length(yr_fe_indx)))
  spatial_field <- array(0, dim = c(nrow(grid_locs), samples, length(yr_fe_indx)))
  proj_matrix   <- INLA::inla.spde.make.A(mesh, loc = as.matrix(grid_locs))

  # loop over all years; do MCMC projections for that year
  for (yr in seq_along(yr_fe_indx)) {
   indx <- which(iset$spatial.field.group == 1) 
      for (i in seq_len(samples)) {
      if (yr > 1) {
        year_diff <- inla.mcmc[[i]]$latent[yr_fe_indx[[yr]]]
      } else {
        year_diff <- 0
      }

      # constant covariate effects across years (currently hard coded...)
      fixed_covariate_effects <- 
        pred_grid[, "depth_scaled", drop = TRUE] * inla.mcmc[[i]]$latent[depth_fe_indx1] +
        pred_grid[, "depth_scaled2", drop = TRUE] * inla.mcmc[[i]]$latent[depth_fe_indx2]

      # spatial temporal 
      spatial_field[, i, yr] <-
         as.numeric(proj_matrix %*% inla.mcmc[[i]]$latent[re_indx][indx]) +
         #inla.mcmc[[i]]$latent[intercept_indx] + 
         inla.mcmc[[i]]$latent[yr_fe_indx[[1]]] + # base year
         year_diff # 0 for the base year, diff otherwise

      # model predictions (current covariates are hard coded)   
      latent_grid[, i, yr] <- spatial_field[, i, yr] + fixed_covariate_effects
   }
  }
  list(latent_grid = latent_grid, depth_effects = depth_effects, spatial_field = spatial_field)
}

@@

```{r posterior-draws}
#pg_one_year <- filter(pg, year == min(pg$year))
projected <- predict_inla_sens(obj = obj, pred_grid = pg, samples = 10)
```

Plot values on the grid:

```{r prediction-grid-calcs}
attributes(projected)
pg_one_year <- filter(pg, year == min(pg$year))
attributes(projected$latent_grid)
# out <- stats::plogis(p_bin) * exp(p_pos)
out_med <- apply(projected$latent_grid, c(1,3), median)
# out_bin <- stats::plogis(p_bin)
# out_bin <- apply(out_bin, c(1, 3), median)
pg_all <- bind_rows(replicate(dim(projected$latent_grid)[3], pg, simplify = FALSE))
out_long <- reshape2::melt(out_med)
pg_all$time <- out_long$Var2
pg_all$pred <- out_long$value

# pg$year_k <- pg$year - min(pg$year) + 1
# count_k <- unique(year_k)

```{r prediction-grid-calcs}
attributes(projected$spatial_field)
# out <- stats::plogis(p_bin) * exp(p_pos)
out_med_sp <- apply(projected$spatial_field, c(1,3), median)
out_long_sp <- reshape2::melt(out_med_sp)
pg_all$sp_field <- out_long_sp$value
```

@@

make_plot <- function(dat, fill = "sqrt(pred)") {
  g <- ggplot(dat, aes(X * 10, Y * 10)) +
    geom_raster(aes_string(fill = fill)) +
    scale_fill_viridis_c(option = "C") +
    geom_polygon(
      data = coast, aes_string(x = "X", y = "Y", group = "PID"),
      fill = "grey80"
    ) +
    geom_point(
      data = filter(d, present == 1),
      aes(x = X * 10, y = Y * 10, size = density), inherit.aes = FALSE,
      pch = 21, col = "white", alpha = 0.3
    ) +
    geom_point(
      data = filter(d, present == 0),
      aes(x = X * 10, y = Y * 10), inherit.aes = FALSE,
      pch = 4, col = "white", alpha = 0.2
    )

@@

make_plot(pg_all, "pred") + facet_wrap(~ time) +
  scale_fill_gradient2(
    midpoint = median(pg_all$pred),
    low = scales::muted("blue"), 
    mid = "white",
    high = scales::muted("red"))

median(pg_all$sp_field)
make_plot(pg_all, "sp_field") + facet_wrap(~ time) +
  scale_fill_gradient2(
    midpoint = median(pg_all$sp_field),
    high = scales::muted("blue"), 
    mid = "white",
    low = scales::muted("red"))
